---
title: |
  | Computer class Module 2: Ecophysiology
date: "`r format(Sys.time(), '%d %B, %Y')`"
author: | 
  | Moi Exposito-Alonso
output: 
  html_document:
   # pdf_document:
   #  toc: yes
   #  latex_engine: xelatex
fontsize: 10pt
mainfont: Times New Roman
indent: true
editor_options: 
  chunk_output_type: console
  preview: viewer
---


# Downloading the data

Download the repository containing this file and data with the command below or simply go to the website and download it directly: https://github.com/MoisesExpositoAlonso/plantclimatelabgit .

```{bash, eval=FALSE}
# SUBSTITUTE THE PATH BELOW FOR YOUR WORKING DIRECTORY
git clone https://github.com/MoisesExpositoAlonso/plantclimatelabgit path/to/the/working/directory
```

Because we already worked with this repository, if you have git installed you can also get the latest using "git pull" command.


```{bash, eval=FALSE}
# SUBSTITUTE THE PATH BELOW FOR YOUR WORKING DIRECTORY
git pull https://github.com/MoisesExpositoAlonso/plantclimatelabgit path/to/the/working/directory
```

# Log your data from the field trip

For reference, the data we gathered is herre:
Leaf traits
https://docs.google.com/spreadsheets/d/1eMJt2Tv-KLKORX2jOT8uhuhdchrMEo02pYtn1zxuzDA/edit?usp=sharing
Additional traits
https://docs.google.com/spreadsheets/d/1_A6K7PmIRX0ZXnnT4RPthZeC_LT4xO_qRruTbJLr2Ac/edit?usp=sharing

Once finished
File -> Download -> Comma Separated Values (.csv). 

# Consolidate data

Now you should have the next files:
- "Module2-leaftraits.csv"
- "Module2-additionaltraits.csv"
In addition, there are other published datasets
- "Pierce.csv"
- "Estarage.csv"
- "GBIF_California"

The folder also contains THIS template called "MODULE2_TUTORIAL.Rmd", which you can use to do the exercises of this practical.

Finally, a subfolder "leaf_area_pictures" contains the pictures that we took for leaf areas

# Dependencies

The exercises in this practical requires an R packages for Ternary plots (Ternary) and a package to manipulate climate (raster, I already used this to clean some data, so no worries if it does not install).

First set a working directory (the folder that contains the code and the data). This is so R knows where your data is

Install and load packages (libraries of code) that we need for the analyses

```{r load libraries, warning=FALSE}

# SUBSTITUTE THIS FOR PATH TO FOLDER WITH DATA:
# working_directory = "path/to/the/working/directory"
# setwd(working_directory)

#to install the packages used for the analysis
# install.packages("raster")
# install.packages("Ternary")

#to load these same packages into R
library(Ternary)
library(raster)

```

# Load the data we gathered from the trip

We quantified data ourselves. The spreadsheet has as input of three traits that are used to quantify composits:

LA leaf area (mm2) [single-sided leaf area with petiole]
LFW leaf fresh weight (mg) [saturated, turgid fresh weight]
LDW leaf dry weight (mg)  [after oven drying to constant weight]	LSI leaf succulence index (g water dm-2)
LWC leaf water content (% of fresh weight)
LMA leaf mass per area (g m-2)
LA leaf area (mm2)
LDMC leaf dry matter content (%)
SLA specific leaf area (mm2 mg-1)

```{r load practical data}

leaftraits = read.csv("Module2-leaftraits.csv",header = T)
leaftraits$species.binomial
othertraits = read.csv("Module2-additionaltraits.csv",header = T)
othertraits$species.binomial

mergedtraits<-
  merge(leaftraits, 
        othertraits,
        by='species.binomial')
```


# Load the data publically available

All the plants from California from GBIF.org

```{r}
# calplants = read.csv("observations-226188.csv")
calplants = read.csv("GBIF_California.csv",sep = '\t')
# calplants = read.csv("../plantclimatelabgit/module1/")
head(calplants)
calplantsclim = read.csv("GBIF_California_clim.csv",sep = '\t')

```


The Pierce method comes with an example dataset (https://besjournals-onlinelibrary-wiley-com.stanford.idm.oclc.org/doi/10.1111/1365-2435.12722
)

You can seem more public species traits here:
http://www.try-db.org/TryWeb/Data.php


```{r load TRY data}
try = read.csv("Pierce.csv",header = T,fill = T)
head(try)
```

Estarague (http://dx.doi.org/10.1093/aob/mcab149) applied the CSR to leaf traits
of the Arabidopsis thaliana model plant.


```{r load Ara data}
esta = read.csv("Estarague.csv",header = T,fill = T)
head(esta)
```


# Question 1: Plotting the global CSR. Where are the different ecosystems of california?

The CSR is a complicated plot, for that we have a package Ternary that makes these
ternary plots. Here is an example with all the plants in the Pierce example spreadsheet. 
You can color the dots to ask whether different habitats have different CSRs.

```{r}
# install.packages("Ternary")
library(Ternary)

TernaryPlot(atip = "C",
            btip = "S",
            ctip = "R"
)
TernaryPoints(try[, c("C....", "S....", "R....")],
              pch = 16 
              )

```

# Question 2: What are the main axis of leaf variation in our dataset?

We quantified several leaf traits today and over the last weeks.
Three traits can be visualized in a 3d plot, but often we want to visualize these
using a Principal Component Analysis. This analysis tries to transform your traits
into new traits which are sorted by how much variation they explain.


Using Leaf area, Leaf dry mass per leaf fresh mass, Specific leaf area (leaf area per dry mass), 
and height, what do you see

```{r}
# Do not run next lines, just mock data to test code
testcode=T
if(testcode){
  mergedtraits$LA<-rnorm(nrow(mergedtraits))
  mergedtraits$LDMC<-rnorm(nrow(mergedtraits))
  mergedtraits$LMA<-rnorm(nrow(mergedtraits))
  mergedtraits$C..............<-rnorm(nrow(mergedtraits))
  mergedtraits$S....<-rnorm(nrow(mergedtraits))
  mergedtraits$R..............<-rnorm(nrow(mergedtraits))
}

```

```{r}
# Subset table for traits
topca<-mergedtraits[,c("LA","LDMC","LMA","height")]
topca<-na.omit(topca)

# run PCA
pca<-prcomp(topca,scale=T)
plot(pca$x)
```

# Question 3: Where are our species in the CSR scale? Is there any relationship between CSR and location of origin?

From the spreadsheet we have generated, you can plot a ternary plot using
the same techniques as above. You can also check how the plants in our trip
compare to the global spectrum. 

To do this, you will have to only use the averages per species. 


```{r}
# Select the average row
# tocsr<-subset(mergedtraits,notes.x=="Mean" ) # do not run
tocsr<-mergedtraits
# Subset table for traits
tocsr<-tocsr[,c( "C..............","S.............." ,"R.............." )]
tocsr<-apply(tocsr,2,as.numeric)
tocsr<-na.omit(tocsr)
# Plot
TernaryPlot(atip = "C",
            btip = "S",
            ctip = "R"
)
TernaryPoints(tocsr,
              pch = 16 
              )

```

# Question 4: Is there intra-species variation in the CSR? how big?

In our trip we observed that some individuals varied in leaves, both within
and between individuals. It may be interesting so see what is the magnitude of this difference

```{r}
hist(as.numeric(esta$LA),xlab="Leaf area [LA]",main="")

```


```{r}
TernaryPlot(atip = "C",
            btip = "S",
            ctip = "R"
)
tocsr<-esta[,c("C","S","R")]
tocsr<-na.omit(tocsr)
TernaryPoints(tocsr,
              pch = 16 
              )
```

# Question 5: Is there any relationship between CSR and climate?

GBIF_California_clim.csv contains the bioclimatic variables for all the GPS locations 
Estarague.csv already also contains the locations of Arabidopsis ecotypes with the climate extracted

Some more information about the climate
https://github.com/MoisesExpositoAlonso/rbioclim
```{r}
# you can do a correlation, fill in this
```

# Question 6: Can simple recurrent equations lead to CSR coexistence?

Go to the module lecture slides
https://docs.google.com/presentation/d/1mAV-e8GAncHZ3Joz7yRdhUlK0IKUsxF2gsWLFAH0VJc/edit?usp=sharing

And try to recode the equations from slides 14. Some tips below on how to start thinking about this.
Just assume 2 species with abundance N1 and N2 for simplicity
And 2 resources R1 R2
Write recursive functions
Give example starting parameters and run the simulation.

What conditions lead to one species to win?
What conditions lead to coexistence?
What combination of parameters you have to choose?

```{r}
N1<-10
N2<-10
f1<-function(Ri){}
f2<-function(Ri){}
m1<-0.2
m2<-0.3
R1<-1
R2<-1
a1<-0.5
S1<-function(){N1*0.5}
S2<-function(){N2*0.8}
alpha<-0.2

dndt<-function(Ni,fi,R1,R2,mi,alpha){
  # write the equation here
}
drdt<-function(Ri,ai,N1,N2,f1,f2){
  # write the equation here
}

generations=50
recordN1<-c()
recordN2<-c()
for(i in 1:generations){
  N1<-dndt(N1,f1,R1,R2,m1,alpha)+N1
  N2<-dndt(N2,f2,R1,R2,m2,alpha)+N1
  R1<-drdt(R1,a1,N1,N2,f1,f2)+R1
  R2<-drdt(R2,a2,N1,N2,f1,f2)+R2
  recordN1<-append(recordN1,N1)
  recordN2<-append(recordN2,N2)
}
```

