---
title: |
  | Computer class Module 1: Plant Biodiversity
date: "`r format(Sys.time(), '%d %B, %Y')`"
author: | 
  | Moi Exposito-Alonso
output: 
  html_document:
   # pdf_document:
   #  toc: yes
   #  latex_engine: xelatex
fontsize: 10pt
mainfont: Times New Roman
indent: true
editor_options: 
  chunk_output_type: console
  preview: viewer
---


# Downloading the data

Download the Jasper Ridge Data: "JR_herbarium.csv" and "JR_observations-218315.csv"

Download the Google Drive transect data as a .csv.

To get both files:
In Google Drive go to each of the tabs (redwood and grassland) and go to
File -> Download -> Comma Separated Values (.csv). Do it for each

You should now have 4 files:
- "Module1-transects - redwood.csv"
- "Module1-transects - grassland.csv"
- "JR_herbarium.csv"
- "JR_observations-218315.csv"


Make a folder in your computer that contains all of these files and
MODULE1_TUTORIAL.Rmd

# Dependencies

Set a working directory (the folder that contains the code and the data). This is so R knows where your data is

Install and load packages (libraries of code) that we need for the analyses

```{r load libraries}

# SUBSTITUTE THIS FOR PATH TO FOLDER WITH DATA:
# working_directory = "path/to/the/working/directory"
# setwd(working_directory)

#to install the packages used for the analysis
# install.packages("sads")
# install.packages("sars")

#to load these same packages into R
library(sads)
library(sars)

```

# Load the data we gathered from the transects

```{r load practical data}

redwood_transect = read.csv("Module1-transects - transect1.csv")
grassland_transect = read.csv("Module1-transects - transect2.csv")

```

# Load the data publically available

This is a list of plant species recorded at Jasper Ridge (Herbarium)

```{r}
JR_herb = read.csv("JR_herbarium.csv")
herb_species = JR_herb$Species

```

We downloaded all the data from iNaturalist observations at Jasper Ridge from this link:
https://www.inaturalist.org/observations/110309697
This data contains unique ids, a temporal footprint, use information, and species ID

```{r load iNat data}

JR_data = read.csv("JR_observations-218315.csv")
JR_species = JR_data$scientific_name

#These lines filter the JR data to only contain identified species of plants (from herbarium data)
rows_with_identified_species = JR_species %in% herb_species
JR_data_quality = subset(JR_data, rows_with_identified_species)
```



# Question 1: Is the data from iNaturalist already capturing most species in JR?

Rarefaction curve. What information is needed to make this curve? 

Open the iNaturalist data from Jasper Ridge. What information in this sheet do you need to make this curve?

Describe how to make this curve using the Jasper Ridge Data if you were to draw it out:



```{r}
sheet_size = nrow(JR_data_quality)
species_names = JR_data_quality$scientific_name
unique_species_names = unique(species_names)
num_species = length(unique_species_names)

#a vector the same size as the number of species with all false entries
species_recorded = rep(FALSE, num_species)
#a vector the same size as the length of the iNaturalist sheet with all NA entries
n_species_overEffort = rep(NA, sheet_size)

total_speciesNum = 0

#Going through the data sheet line by line to get the relevant information for the curve
for(observation in 1:sheet_size) {
  species_name = species_names[observation]
  #Matches where the species_name found on the current row is on unique_species_names 
  species_index = match(species_name,unique_species_names)
  
  #What does this do? What is species_recorded for?
  if(!(species_recorded[species_index])) {
    species_recorded[species_index] = TRUE
    total_speciesNum = total_speciesNum + 1
  }
  
  #How does this help make the curve?
  n_species_overEffort[observation] = total_speciesNum
}

plot(n_species_overEffort, ylab="Number of Species",xlab="Effort (Number of Observations)")

```

What is this code doing? What is stored in n_species_overEffort?



Now we can answer: Is the data from iNaturalist already capturing most species in JR? Why?



# Question 2: Did we sample enough in our transect?

What is different between the way that we collected data for the transect and the way that the iNaturalist data is collected? What would you say the Effort is?



Write code that uses the transect data to make a rarefaction curve. Think about what has to be different considering the differences in data collection

```{r}

```

Compare the 2 transects.

Now that you've seen the rarefaction curve for the transects: Did we sample enough in our transect? Why?


# Question 3: How different species abundance curve models fit the two ecosystems?

Species Abundance Curve. What information is needed to make this curve? 


Open the iNaturalist data from Jasper Ridge. What information in this sheet do you need to make this curve?


Describe how to make this curve using the Jasper Ridge Data if you were to draw it out:




```{r}
sheet_size = nrow(JR_data_quality)
species_names = JR_data_quality$scientific_name
unique_species_names = unique(species_names)
num_species = length(unique_species_names)

#a vector the same size as the number of species with all 0 entries
species_abundance = rep(0, num_species)

#Going through the data sheet line by line to get the relevant information for the curve
for(observation in 1:sheet_size) {
  species_name = species_names[observation]
  #Matches where the species_name found on the current row is on unique_species_names
  species_index = match(species_name,unique_species_names)
  
  #How does this help make the curve?
  species_abundance[species_index] = species_abundance[species_index] + 1
}

#Rank Abundance Distribution (rad) objects. Look at species_abundance.rad
species_abundance.rad = rad(species_abundance)
plot(species_abundance.rad, ylab="Number of individuals")


```

What is this code doing? What is stored in species_abundance?



The following code fits models to the data and plots it:

```{r}
#fitsad fits different models to the species abundance data
#radpred creates a rank abundance table from on the model


log_normal = fitsad(species_abundance,"lnorm")
log_normal.rad = radpred(log_normal)

broken_stick = fitsad(species_abundance,"bs")
broken_stick.rad = radpred(broken_stick)

geometric = fitsad(species_abundance,"geom")
geometric.rad = radpred(geometric)

log_series = fitsad(species_abundance,"ls")
log_series.rad = radpred(log_series)

plot(species_abundance.rad, ylab="Number of individuals")
lines(log_normal.rad, ylab="Number of individuals", col="blue")
lines(broken_stick.rad, ylab="Number of individuals", col="green")
lines(geometric.rad, ylab="Number of individuals", col="red")
lines(log_series.rad, ylab="Number of individuals", col="magenta")
legend("topright",c("log normal", "broken stick", "geometric", "log series"), lty=1, col=c("blue", "red", "green", "magenta"))


#This table helps us choose which model has the best fit
AICtab(log_normal, broken_stick, geometric, log_series, base = TRUE)
```

Using the data from both transects, write code to create your own species abundance curve models:


```{r}

```

Compare the 2 transects. 

Which model is a better fit?


# Question 4: Is there a species area curve in JR?

What is required to make a species area curve? 



How is the following code doing that? Describe it in detail


```{r}
longitudes = JR_data$longitude
latitudes = JR_data$latitude

min_x = min(longitudes)
max_x = max(longitudes)

min_y = min(latitudes)
max_y = max(latitudes)

increment_fraction = 0.05
increment_x = (max_x-min_x)*increment_fraction
increment_y =  (max_y-min_y)*increment_fraction

current_x = min_x + increment_x
current_y = min_y + increment_y

#empty vectors
area = c()
Species = c()

current_area = 0
num_species_in_area_range = 0
message("Area: ",current_area,", species: ",num_species_in_area_range)

while(current_x < max_x) {
  while(current_y < max_y) {
    
    species_in_area_range = subset(JR_data_quality, longitude<=current_x, latitude<=current_y)
    unique_species_in_area_range = unique(species_in_area_range$scientific_name)
    num_species_in_area_range = length(unique_species_in_area_range)
    Species = c(Species,num_species_in_area_range)
    
    area_range_x = diff(range(species_in_area_range$longitude))
    area_range_y = diff(range(species_in_area_range$latitude))
    
    current_area = area_range_x * area_range_y
    if(is.infinite(current_area)) {
      current_area = 0
    }
    area<-c(area, current_area)
    message("Area: ",current_area,", species: ",num_species_in_area_range)
    
    # increase the area for the next loop
    current_x = current_x + increment_x
    current_y = current_y + increment_y
  }
}

species_area_table = data.frame(a=area,S=Species)
#sar_pover creates a SAR curve
SAR_curve = sars::sar_power(species_area_table)
plot(SAR_curve)
```


Do you think that a SAR with the transect data could give meaningful results? Why?


# Question 5: What will happen to Jasper Ridge with masticator treatment?

If we remove a part of the area, how many species are lost? 

Remove a part of the area and calculate the amount of species lost

```{r}

```

